---
title: "Example 2: Fetch data from Statistics Estonia with R"
author: "Nicolas"
date: "2025-04-17"
institute: "Eesti Pank"
format:
  revealjs:
    # output-file: presentation_for_distribution_fetching.html
    encoding: "UTF-8"
    incremental: true
    preview-links: true
    scroll-view: false
    theme: simple
    smaller: true
    slide-number: true
    font-size: 0.6em
    transition: slide
    toc: false
    toc-depth: 2
    chalkboard: false
    embed-resources: false
  docx:
    toc: true
    number-sections: true
    highlight-style: github
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
  beamer:
    slide-level: 2
    output-file: example2_beamer.pdf
editor_options:
  chunk_output_type: console
---

## Setup R

::: fragment
```{r}
library(quarto)     # for compiling Quarto presentations and documents
library(pxweb) # query and fetch data from Statistics Estonia
library(tidyverse)  # dplyr, ggplot2, readr, etc.
library(here)  # path settings
```
:::

## Accessing Statistics Estonia Data with `pxweb`

-   [Statistics Estonia](https://www.stat.ee/en) is **not yet part of DBnomics**
-   There is **no dedicated R package** for [andmed.stat.ee](https://andmed.stat.ee/en/stat)
-   **However**, you can access the data using the `pxweb` R package
-   It is based on the **PX-WEB API**, also used by many Nordic statistical agencies

------------------------------------------------------------------------

### Data Access Options

-   **Interactive Access**\
    Browse tables and build queries via the Statistics Estonia website in the R console

-   **Direct Programmatic Query**\
    Use the API endpoint (a *URL*) and supply **key-value pairs** to specify:

    -   Variables (e.g., indicators, time, sector)
    -   Values or selections for each variable

## Find and download data interactively

```{r}
#| eval: false
pxweb_interactive() # All available databases
pxweb_interactive("https://andmed.stat.ee/api/v1/en/stat") # Statistics Estonia
```

``` r
#| eval: false
============================================================================================================================
 R PXWEB: Content of 'andmed.stat.ee'
          at '/api/v1/en/stat'
============================================================================================================================
 [ 1 ] : Environment
 [ 2 ] : Economy
 [ 3 ] : Population
 [ 4 ] : Social life
 [ 5 ] : Multidomain statistics
 [ 6 ] : Population and Housing Census
 [ 7 ] : Discontinued datasets
============================================================================================================================
Enter your choice:
('esc' = Quit, 'b' = Back, 'i' = Show id)
1:
```

## GET call to PXWEB API

 - Example dataset:  PA111 -  AVERAGE MONTHLY GROSS WAGES (SALARIES)

::: fragment
```{r}
pa111_meta <- pxweb_get(
  url = "https://andmed.stat.ee/api/v1/en/stat/PA111")
pa111_meta
```

The object is just metadata, no data yet!

*How to identify key-value pairs?*

-   Use either:
    1.  Programmatically via the metadata object
    2.  From the Statistics Estonia web interface – click **"API query for this table"**
:::

## Key-value pairs

-   Values for the first key (Näitaja)

::: fragment
```{r}
pa111_meta$variables[[1]]
```
:::

## Fetch data

-   Set: Key-value pair (as R list object) & Url to the database
  -   Wildcard operator `*` to select all values

::: fragment
```{r}
px_query_list1 <- list(
  "Näitaja" = c("GR_W_D1", "GR_W_D5", "GR_W_D9"),
  "Tegevusala" = "*",
  "Vaatlusperiood" = "*"
)

pa111 <- pxweb_get(
  url = "https://andmed.stat.ee/api/v1/en/stat/PA111",
  query = px_query_list1
)
```
:::

-   We get a list object with the data

::: fragment
```{r}
pa111
class(pa111)
```
:::

## Convert the pxweb list object to a data frame

```{r}
# Convert to data frame
pa111.df <- as.data.frame(pa111, column.name.type = "text", variable.value.type = "text")
pa111.code <- as.data.frame(pa111, column.name.type = "code", variable.value.type = "code")
# Optional: Combine text and code in one data frame
pa111.df$Näitaja <- pa111.code$Näitaja
```

```{r}
head(pa111.df, n=3)
```

```{r}
#| echo: false
selected_sectors <- c("Manufacturing", "Construction", "Information and communication", "Real estate activities", "Total - all activities")
# Clean and reshape
pa111_plotdata <- pa111.df %>%
  mutate(
    quarter = yq(str_replace(`Reference period`, " ", "-")),
    Indicator = recode(
      Näitaja,
      "GR_W_D1" = "D1",
      "GR_W_D5" = "Median",
      "GR_W_D9" = "D9"
    )
  ) %>%
  filter(
    `Economic activity` %in% selected_sectors,
    Indicator %in% c("D1", "Median", "D9")
  ) %>%
  select(quarter, `Economic activity`, Indicator, value = `PA111: AVERAGE MONTHLY GROSS WAGES (SALARIES), MEDIAN, DECILES AND NUMBER OF EMPLOYEES`) %>%
  pivot_wider(names_from = Indicator, values_from = value)
```

```{r}
#| echo: false
overall_median <- pa111_plotdata %>%
  filter(`Economic activity` == "Total - all activities") %>%
  select(quarter, total_median = Median)

pa111_plotdata <- pa111_plotdata %>%
  filter(`Economic activity` != "Total - all activities") %>%
  left_join(overall_median, by = "quarter")
```

```{r}
#| echo: false
meta <- pa111$metadata[[1]]
title <- meta$label
source <- meta$source

source_note <- glue::glue(
  "Source: {source}\n\"{title}\""
)
```

------------------------------------------------------------------------

```{r}
ggplot(pa111_plotdata, aes(x = quarter)) +
  geom_ribbon(aes(ymin = D1, ymax = D9), fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = Median), color = "darkblue", linewidth = 1) +
  geom_line(aes(y = total_median), color = "red", linetype = "dashed", linewidth = 0.7) +
  facet_wrap(~`Economic activity`, ncol = 2) +
  labs(
    title = "Wage Distribution by Sector (1st to 9th Decile, Median)",
    subtitle = "Red dashed line shows overall median (Total - all activities)",
     caption = source_note,
    x = "",
    y = "Gross Nominal Monthly Wages (EUR)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Rendering All Formats from R or Bash

To render the document to **all output formats** (docx, Beamer, Reveal.js): run programmatically from another script or the R console:

```{r}
#| eval: false
quarto::quarto_render(here("teaching_materials", "example2_statistics_estonia.qmd"), output_format = "all")
```

or from the command line:

``` bash
quarto render example2_statistics_estonia.qmd
```

## End of session
