---
title: "Example 2: Fetch from Statistics Estonia with R"
author: "Nicolas"
date: "2025-04-17"
institute: "Eesti Pank"
format:
  revealjs:
    # output-file: presentation_for_distribution_fetching.html
    encoding: "UTF-8"
    incremental: true
    preview-links: true
    scroll-view: false
    theme: simple
    smaller: true
    slide-number: true
    font-size: 0.6em
    transition: slide
    toc: true
    toc-depth: 2
    chalkboard: false
    embed-resources: false
  docx:
    toc: true
    number-sections: true
    highlight-style: github
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
  beamer:
    slide-level: 2
    output-file: example2_beamer.pdf
editor_options:
  chunk_output_type: console
---

```{r}
library(quarto)     # for compiling Quarto presentations
library(pxweb)
library(tidyverse)  # dplyr, ggplot2, readr, etc.
library(here)  # path
```


## Accessing Statistics Estonia Data with `pxweb`

- [Statistics Estonia](https://www.stat.ee/en) is **not yet part of DBnomics**
- No dedicated R package for [andmed.stat.ee](https://andmed.stat.ee/en/stat)
- But: you can access data using the `pxweb` R package
- Based on the **PX-WEB API**, which is also used by many Nordic stats agencies

---

### 🔍 How to Identify Variables

## How to Identify Query Variables

- 🧠 Use either:
  1. Programmatically via the metadata (`pxweb_get()`, `pxweb_get_metadata()`)
  2. From the Statistics Estonia web interface – click **"API query for this table"**

```{r}
pa111_meta <- pxweb_get(
  url = "https://andmed.stat.ee/api/v1/en/stat/PA111")
pa111_meta
pa111_meta$variables[[1]]
```

--- 


```{r}
px_query_list1 <- list(
  "Näitaja" = c("GR_W_D1", "GR_W_D5", "GR_W_D9"),
  "Tegevusala" = "*",
  "Vaatlusperiood" = "*"
)

pa111 <- pxweb_get(
  url = "https://andmed.stat.ee/api/v1/en/stat/PA111",
  query = px_query_list1
)

# Convert to data frame
pa111.df <- as.data.frame(pa111, column.name.type = "text", variable.value.type = "text")
pa111.code <- as.data.frame(pa111, column.name.type = "code", variable.value.type = "code")
pa111.df$Näitaja <- pa111.code$Näitaja
```

```{r}
selected_sectors <- c("Manufacturing", "Construction", "Information and communication", "Real estate activities", "Total - all activities")
# Clean and reshape
pa111_plotdata <- pa111.df %>%
  mutate(
    quarter = yq(str_replace(`Reference period`, " ", "-")),
    Indicator = recode(
      Näitaja,
      "GR_W_D1" = "D1",
      "GR_W_D5" = "Median",
      "GR_W_D9" = "D9"
    )
  ) %>%
  filter(
    `Economic activity` %in% selected_sectors,
    Indicator %in% c("D1", "Median", "D9")
  ) %>%
  select(quarter, `Economic activity`, Indicator, value = `PA111: AVERAGE MONTHLY GROSS WAGES (SALARIES), MEDIAN, DECILES AND NUMBER OF EMPLOYEES`) %>%
  pivot_wider(names_from = Indicator, values_from = value)
```

---

```{r}
overall_median <- pa111_plotdata %>%
  filter(`Economic activity` == "Total - all activities") %>%
  select(quarter, total_median = Median)
```

```{r}
pa111_plotdata <- pa111_plotdata %>%
  filter(`Economic activity` != "Total - all activities") %>%
  left_join(overall_median, by = "quarter")
```

---

```{r}
ggplot(pa111_plotdata, aes(x = quarter)) +
  geom_ribbon(aes(ymin = D1, ymax = D9), fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = Median), color = "darkblue", linewidth = 1) +
  geom_line(aes(y = total_median), color = "red", linetype = "dashed", linewidth = 0.7) +
  facet_wrap(~`Economic activity`, ncol = 2) +
  labs(
    title = "Wage Distribution by Sector (1st to 9th Decile, Median)",
    subtitle = "Red dashed line shows overall median (Total - all activities)",
    x = "Quarter",
    y = "Gross Nominal Monthly Wages (EUR)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Rendering All Formats from R or Bash

T render the document to **all output formats** (docx, Beamer, Reveal.js): run  programmatically from another script or the R console:

```{r}
#| eval: false
quarto::quarto_render(here("teaching_materials", "example2_statistics_estonia.qmd"), output_format = "all")
```

or from the command line: 

```bash
quarto render example2_statistics_estonia.qmd
```
