---
title: "Fetching and Visualizing Official Statistics with R"
author: "Nicolas"
date: "2025-04-17"
institute: "Eesti Pank"
format:
   revealjs:
    output-file: presentation_for_distribution_fetching.html
    encoding: "UTF-8"
    incremental: true
    preview-links: true
    scroll-view: false
    theme: simple
    smaller: true
    slide-number: true
    font-size: 0.6em
    transition: slide
    toc: true
    toc-depth: 2
    chalkboard: false
    embed-resources: false
editor_options:
  chunk_output_type: console
---

# Interfaces to Official Statistics

- Packages or set of classes and methods to read data and metadata documents through exchange frameworks
  - Use [R](https://www.r-project.org/) (or [Python](https://www.python.org/)) packages to read data from APIs, databases, and web pages
    - Individual packages:
      - [eurostat](https://cran.r-project.org/package=eurostat): Access data from Eurostat
      - [OECD](https://cran.r-project.org/package=OECD): Access data from the OECD API
  - General-purpose packages:
    - [rsdmx](https://cran.r-project.org/package=rsdmx): Direct access to SDMX endpoints (Eurostat, ECB, OECD, IMF, etc.)

---

  - Interface standards:
    - [SDMX](https://sdmx.org/?page_id=5008): Statistical Data and Metadata Exchange format
    - [pxweb](https://cran.r-project.org/package=pxweb): Access to data sources using the PX-Web API (e.g. Statistics Sweden, Statistics Estonia)

---

### What is SDMX?

- **S**tatistical **D**ata and **M**etadata e**X**change
  - International standard for exchanging statistical data
  - Initiated by 7 international organizations: BIS, ECB, Eurostat, IMF, OECD, UN, World Bank
  - Direct access to official statistics at the source
  - Standardized data structures and metadata across providers

## How to fetch data (using rsdmx)

- rsdmx R package for SDMX data access

::: {.fragment}

```{r}
#| eval: false
#| echo: true
install.packages("rsdmx")
library(rsdmx)
```

:::

---

### Packages used in this tutorial


  - Fetching data (`rsdmx`)
  - Data wrangling and transformation (`tidyverse`, `lubridate`)
  - Visualization (`ggplot2`, `plotly`)
  - Tabular summaries (`gt`)
  - Building this presentation (`quarto`)

::: {.fragment}

```{r}
#| message: false
#| warning: false
library(quarto)       # for compiling Quarto presentations
library(rsdmx)        # for accessing SDMX data directly
library(pxweb)        # for PX-Web API
library(tidyverse)    # dplyr, ggplot2, readr, etc.
library(lubridate)    # for date handling
library(plotly)       # interactive visualizations
library(gt)           # pretty tables
library(countrycode)  # country code conversion
```

:::

---

### Exploring SDMX Providers

::: {.fragment}

```{r}
# List available built-in providers
providers <- getSDMXServiceProviders()
provider_df <- as.data.frame(providers)
provider_df[, c("agencyId", "name")] |> head(12)
```

:::

---

### Example: Fetch Unemployment Data

   - Assume we know exactly the series key we want to fetch
     - Unemployment rate, ILO definition, total, Estonia, from Eurostat

::: {.fragment}

```{r}
#| cache: true
# Fetch unemployment data from Eurostat
unemp_sdmx <- readSDMX(
  providerId = "ESTAT",
  resource = "data",
  flowRef = "ei_lmhr_m",
  key = "M.PC_ACT.SA.LM-UN-T-TOT.EE",
  start = 2000,
  end = 2025
)

# Convert to data.frame
unemp <- as.data.frame(unemp_sdmx)
```

:::

---


```{r}
glimpse(unemp)
colnames(unemp)
```

---

```{r}
# Standardize the data
unemp <- unemp |>
  mutate(
    period = ym(obsTime),
    value = as.numeric(obsValue),
    provider_code = "Eurostat",
    dataset_code = "ei_lmhr_m",
    series_code = "M.PC_ACT.SA.LM-UN-T-TOT.EE"
  )

# Extract metadata
(source_name <- unique(unemp$dataset_code))
(provider_code <- unique(unemp$provider_code))
(country_name <- "Estonia")
(series_id <- unique(unemp$series_code))
```

---

```{r}
#| echo: true
#| output-location: slide
# Plot the data
p1 <- ggplot(unemp, aes(x = period, y = value)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    title = paste("Unemployment Rate in", country_name),
    subtitle = paste("Monthly, seasonally adjusted —", provider_code),
    x = "Date", y = "Percent",
    caption = paste("Source:", provider_code, "| Dataset:", source_name, "| ID:", series_id)
  ) +
  theme_minimal()
p1
```

---

### Interactive plot

```{r}
ggplotly(p1)
```

---

```{r}
#| echo: true
#| output-location: slide
source(here::here("scripts", "theme_ep.R"))
ggplot(unemp, aes(x = period, y = value)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 0, color = "#000", linewidth = 0.5) +
  scale_x_date(
    date_breaks = "2 year",
    date_labels = "%Y"
  ) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "%"),
    name = NULL
  ) +
  labs(
    title = paste("Unemployment Rate in", country_name),
    subtitle = paste("Monthly, seasonally adjusted"),
    x = NULL,
    caption = paste(
      "Source:",
      provider_code,
      "| Dataset:",
      source_name,
      "| ID:",
      series_id
    )
  ) +
  theme_ep()
```

---

### How do we find the series key?

- Go to the provider's data portal:
  - [Eurostat Data Browser](https://ec.europa.eu/eurostat/databrowser/)
  - [ECB Data Portal](https://data.ecb.europa.eu/)
  - [OECD Data Explorer](https://data-explorer.oecd.org/)
- Look for SDMX download options or API query builders
- Or explore programmatically with rsdmx

## Fetch data from ECB

 - Example: MFI Interest Rate Statistics from the ECB
    - Estonia mortgage rates (loans to households for house purchase)

::: {.fragment}

```{r}
#| cache: true
# ECB uses direct URL approach
ecb_url <- "https://data-api.ecb.europa.eu/service/data/MIR/M.EE.B.A2C.A.R.A.2250.EUR.N?startPeriod=2005&endPeriod=2025"

mir_ee_sdmx <- readSDMX(ecb_url)
mir_ee <- as.data.frame(mir_ee_sdmx)

# Standardize
mir_ee <- mir_ee |>
  mutate(
    period = ym(obsTime),
    value = as.numeric(obsValue)
  )

head(mir_ee[, c("period", "value", "TITLE")], 3)
```

:::

---

### Wildcarding dimensions

  - To fetch **multiple values** for a dimension (e.g. countries), leave that position empty in the key
    - Example: remove `"EE"` to fetch all countries (`REF_AREA`)

::: {.fragment}

<span style="color:red;">This can take a while</span>

```{r}
#| cache: true
# Wildcard the country dimension (empty between dots)
mir_url <- "https://data-api.ecb.europa.eu/service/data/MIR/M..B..A.R.A..EUR.N?startPeriod=2015&endPeriod=2025"

mir_sdmx <- readSDMX(mir_url)
mir <- as.data.frame(mir_sdmx) |>
  mutate(
    period = ym(obsTime),
    value = as.numeric(obsValue)
  )

unique(mir$REF_AREA)
unique(mir$BS_ITEM)
```

:::

---

### Filter and plot

- Filter Estonia, Latvia and Lithuania
- Keep only selected **BS items** (loan categories)

::: {.fragment}

```{r}
# Filter by BS_ITEM and countries
mir_filtered <- mir %>%
  filter(
    REF_AREA %in% c("EE", "LV", "LT"),
    BS_ITEM %in% c("A2I", "A2C", "A2B", "A2J", "A2A")
  )
```

:::

---

### Plot interest rates by country & type

::: {.fragment}


```{r}
#| echo: true
#| output-location: slide
country_list <- paste(sort(unique(mir_filtered$REF_AREA)), collapse = ", ")
item_list <- paste(unique(mir_filtered$BS_ITEM), collapse = ", ")

caption_text <- paste(
  "Source: ECB MIR Statistics",
  paste0("\nFiltered: REF_AREA in ", country_list, "; BS_ITEM in ", item_list)
)

mir_filtered <- mir_filtered %>%
  mutate(facet_label = paste0(BS_ITEM, ": ", TITLE_COMPL |> str_trunc(50)))

p4 <- ggplot(mir_filtered, aes(x = period, y = value, color = REF_AREA)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ BS_ITEM, ncol = 3) +
  labs(
    title = "Interest Rates for Households and Firms",
    subtitle = "Faceted by Loan Type",
    x = "Date", y = "Percent",
    caption = caption_text
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
p4
```

:::

## Fetch data from OECD

- OECD updated their API in July 2024 (OECD Data Explorer)
- Use direct URL approach with the new API format

::: {.fragment}

```{r}
#| cache: true
# OECD Composite Leading Indicator
oecd_url <- "https://sdmx.oecd.org/public/rest/data/OECD.SDD.STES,DSD_STES@DF_CLI/.M.LI...AA...H?startPeriod=2015-01&endPeriod=2025-12"

cli_sdmx <- readSDMX(oecd_url)
cli <- as.data.frame(cli_sdmx) |>
  mutate(
    period = ym(obsTime),
    value = as.numeric(obsValue)
  )

head(cli[, c("REF_AREA", "period", "value")], 5)
```

:::

---

### Wildcarding and filtering

```{r}
#| warning: false
# Filter for European countries
cli_eu <- cli |>
  filter(REF_AREA %in% c("EU27_2020", "AUT", "EST", "FIN", "DEU", "LVA", "LTU", "POL", "SWE")) |>
  mutate(
    country = case_when(
      REF_AREA == "EU27_2020" ~ "EU27",
      TRUE ~ countrycode(REF_AREA, origin = "iso3c", destination = "country.name")
    )
  )
unique(cli_eu$country)
```

```{r}
#| echo: false
#| warning: false
library(gghighlight)

desired_order <- c("European Union", "Austria", "Germany", "Estonia",
                   "Finland", "Lithuania", "Latvia", "Poland", "Sweden")

p3 <- cli_eu |>
  ggplot() +
  geom_hline(yintercept = 100, linetype = "solid", size = 0.25) +
  geom_point(
    data = cli_eu |> group_by(country) |> slice_max(period),
    aes(x = period, y = value, color = country),
    shape = 16
  ) +
  geom_line(aes(x = period, y = value, color = country)) +
  gghighlight(
    use_direct_label = FALSE,
    unhighlighted_params = list(colour = alpha("grey85", 1))
  ) +
  geom_text(
    data = cli_eu |> group_by(country) |> slice_max(period),
    aes(x = period, y = value, color = country, label = round(value, 1)),
    hjust = -0.5,
    vjust = 0.5,
    size = 2.5,
    fontface = "bold"
  ) +
  scale_x_date(
    date_labels = "%Y",
    date_breaks = "2 years",
    name = NULL
  ) +
  scale_y_continuous(
    breaks = c(96, 98, 100, 102, 104),
    name = NULL
  ) +
  facet_wrap(~ factor(country, levels = desired_order)) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  theme(legend.position = "none") +
  labs(
    title = "Composite Leading Indicator",
    subtitle = "Selected European Countries & EU27",
    caption = "Source: OECD / rsdmx"
  )
```

---

```{r}
#| echo: false
p3
```

## Accessing Statistics Estonia Data with `pxweb`

- [Statistics Estonia](https://www.stat.ee/en) is **not available via SDMX**
  - There is **no dedicated R package** for [andmed.stat.ee](https://andmed.stat.ee/en/stat)
- **However**, you can access the data using the `pxweb` R package
- It is based on the **PX-WEB API**, also used by many Nordic statistical agencies

---

  ### Data Access Options

  - **Interactive Access**
  Browse tables and build queries via the Statistics Estonia website in the R console

- **Direct Programmatic Query**
  Use the API endpoint (a *URL*) and supply **key-value pairs** to specify:
  - Variables (e.g., indicators, time, sector)
- Values or selections for each variable


##   Find and download data interactively

```{r}
#| eval: false
pxweb_interactive() # All available databases
pxweb_interactive("https://andmed.stat.ee/api/v1/en/stat") # Statistics Estonia
```

```r
#| eval: false
============================================================================================================================
  R PXWEB: Content of 'andmed.stat.ee'
at '/api/v1/en/stat'
============================================================================================================================
  [ 1 ] : Environment
[ 2 ] : Economy
[ 3 ] : Population
[ 4 ] : Social life
[ 5 ] : Multidomain statistics
[ 6 ] : Population and Housing Census
[ 7 ] : Discontinued datasets
============================================================================================================================
  Enter your choice:
  ('esc' = Quit, 'b' = Back, 'i' = Show id)
1:
  ```

## GET call to PXWEB API

::: {.fragment}

```{r}
pa111_meta <- pxweb_get(
  url = "https://andmed.stat.ee/api/v1/en/stat/PA111")
pa111_meta
```

The object is just metadata, no data yet!

  *How to identify key-value pairs?*

  - Use either:
  1.  Programmatically via the metadata (`pxweb_get()`)
2.  From the Statistics Estonia web interface – click **"API query for this table"**

  :::

  ## Key-value pairs


  - Values for the first key (Näitaja)

::: {.fragment}

```{r}
pa111_meta$variables[[1]]
```

:::

  ## Fetch data


  -   Set: Key-value pair (as R list object) & Url to the database
-   Wildcard operator `*` to select all values

::: {.fragment}

```{r}
px_query_list1 <- list(
  "Näitaja" = c("GR_W_D1", "GR_W_D5", "GR_W_D9"),
  "Tegevusala" = "*",
  "Vaatlusperiood" = "*"
)

pa111 <- pxweb_get(
  url = "https://andmed.stat.ee/api/v1/en/stat/PA111",
  query = px_query_list1
)
```


:::

  -   We get a list object with the data

::: {.fragment}


```{r}
pa111
class(pa111)
```


:::


## Convert the pxweb list object to a data frame

```{r}
# Convert to data frame
pa111.df <- as.data.frame(pa111, column.name.type = "text", variable.value.type = "text")
pa111.code <- as.data.frame(pa111, column.name.type = "code", variable.value.type = "code")
# Optional: Combine text and code in one data frame
pa111.df$Näitaja <- pa111.code$Näitaja
```

```{r}
head(pa111.df, n=3)
```


```{r}
#| echo: true
selected_sectors <- c("Manufacturing", "Construction", "Information and communication", "Real estate activities", "Total - all activities")
# Clean and reshape
pa111_plotdata <- pa111.df %>%
  mutate(
    quarter = yq(str_replace(`Reference period`, " ", "-")),
    Indicator = recode(
      Näitaja,
      "GR_W_D1" = "D1",
      "GR_W_D5" = "Median",
      "GR_W_D9" = "D9"
    )
  ) %>%
  filter(
    `Economic activity` %in% selected_sectors,
    Indicator %in% c("D1", "Median", "D9")
  ) %>%
  select(quarter, `Economic activity`, Indicator, value = `PA111: AVERAGE MONTHLY GROSS WAGES (SALARIES), MEDIAN, DECILES AND NUMBER OF EMPLOYEES`) %>%
  pivot_wider(names_from = Indicator, values_from = value)
```

```{r}
#| echo: true
overall_median <- pa111_plotdata %>%
  filter(`Economic activity` == "Total - all activities") %>%
  select(quarter, total_median = Median)

pa111_plotdata <- pa111_plotdata %>%
  filter(`Economic activity` != "Total - all activities") %>%
  left_join(overall_median, by = "quarter")
```

```{r}
#| echo: true
meta <- pa111$metadata[[1]]
title <- meta$label
source <- meta$source

source_note <- glue::glue(
  "Source: {source}\n\"{title}\""
)
```


---

```{r}
#| output-location: slide
ggplot(pa111_plotdata, aes(x = quarter)) +
  geom_ribbon(aes(ymin = D1, ymax = D9), fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = Median), color = "darkblue", linewidth = 1) +
  geom_line(aes(y = total_median), color = "red", linetype = "dashed", linewidth = 0.7) +
  facet_wrap(~`Economic activity`, ncol = 2) +
  labs(
    title = "Wage Distribution by Sector (1st to 9th Decile, Median)",
    subtitle = "Red dashed line shows overall median (Total - all activities)",
    caption = source_note,
    x = "",
    y = "Gross Nominal Monthly Wages (EUR)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Summary: rsdmx Patterns

| Pattern | Code Example |
|---------|--------------|
| Built-in provider | `readSDMX(providerId = "ESTAT", resource = "data", flowRef = "dataset", key = "...")` |
| Direct URL | `readSDMX("https://data-api.ecb.europa.eu/service/data/...")` |
| Wildcarding | Leave position empty in key: `"M..SA.."` |
| Get labels | Use `dsd = TRUE`, then `as.data.frame(sdmx, labels = TRUE)` |

## End

